Efficient pedigree recording for fast population genetics simulation

### 2017-12-30 forwards_paper.tex
Tweak time complexity statement per #37

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

To demonstrate the method's flexibility and to test it in different contexts,
we implemented these simulations separately using two different simulation engines:
\begin{enumerate}

    \item
        In C++, using \fwdpp{} library functions and interface code (details below),
        using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
        package based on \fwdpp.
        % $\mu$ and $r$ were the means of Poisson processes,
        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
        % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

@@

Any simulation scheme that records data into tables,
as does Algorithm~\algref{W},
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?  We did some limited testing with the C++
implementation of our simulations, and found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
algorithm itself, which we show in the next section has linear time complexity.

@@

%% simplify run time
Above we have described memory requirements;
how does the computation time required for simplification scale?
Simply because it must process each edge,
the simplification algorithm is linear in the number of edges of the input tree sequence.
This is seen in Figure~\ref{fig:simplify_complexity}X,
where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$.
this is also seen in Figure~\ref{fig:simplify_complexity}X,
showing the time required to simplify progressively larger subsets of a single large tree sequence.

@@

The \msprime{} Tables API is designed to be easily ``plugged'' into any simulation engine.
One simply needs to fill the relevant columns,
and to send them to \msprime{} at regular intervals for simplification.
For example, in the benchmarking simulations above,
we used the \fwdpp{} C++ API \citep{fwdpp}
to fill arrays of simple structures representing the node and edge data.
Doing so requires no modification of the \fwdpp{} code base.
Rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating event.
Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
the node/edge arrays on the C++ side are made visible, without copy, to Python as a multi-column numpy array,
which is processed by the \msprime{} tables API.
Our implementation also uses \texttt{pybind11}
to create a Python package extending \texttt{fwdpy11}
(a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
% (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
% API from a command-line C/C++ application via an embedded Python interpreter.)

@@

%%%%%%%%%%%%%%
% \section*{Methods}

@@

\subsection{Reagents list for the C++ implementation}

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item msprime master branch
\end{enumerate}

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

To demonstrate the method's flexibility and to test it in different contexts,
we implemented these simulations separately using two different simulation engines:

\jda{Note/TODO: the varying references to c++, python, fwdpp, fwdpy11 and
	simupop are probably confusing to a new reader. This is in part because
	we refer to the python/simupop method in various places as one or the
	other; the same goes for c++/fwdpy11/fwdpp. We should decide what to
	call these and make our usage consistent}

\begin{enumerate}

    \item
        In \cpp{}, using \fwdpp{} library functions and interface code (details below),
        using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
        package based on \fwdpp.
        % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
        % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

@@

Any simulation scheme that records data into tables,
as does Algorithm~\algref{W},
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?  We did some limited testing with the \cpp{}
implementation of our simulations, and found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
algorithm itself, which we show in the next section has linear time complexity.

@@

%% simplify run time
Above we have described memory requirements;
how does the computation time required for simplification scale?
Simply because it must process each edge,
% below per discussion
% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
the simplification algorithm is at least linear in the number of edges of the input tree sequence.
Empirically the algorithm is exactly linear,
as seen in Figure~\ref{fig:simplify_complexity}X,
where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$.
this is also seen in Figure~\ref{fig:simplify_complexity}X,
showing the time required to simplify progressively larger subsets of a single large tree sequence.

@@

The \msprime{} Tables API is designed to be easily ``plugged'' into any simulation engine.
One simply needs to fill the relevant columns,
and to send them to \msprime{} at regular intervals for simplification.
For example, in the benchmarking simulations above,
we used the \fwdpp{} \cpp{} API \citep{fwdpp}
to fill arrays of simple structures representing the node and edge data.
Doing so requires no modification of the \fwdpp{} code base.
Rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating event.
Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
which is processed by the \msprime{} tables API.
Our implementation also uses \texttt{pybind11}
to create a Python package extending \texttt{fwdpy11}
(a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
% (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
% API from a command-line C/\cpp{} application via an embedded Python interpreter.)

@@

%%%%%%%%%%%%%%
\section*{Methods}

@@

\subsection{Requirements for the \cpp{} implementation}

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item \msprime{} master branch \jda{which commit hash or short version number
		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
\end{enumerate}
